The Calculus of Linear Constructions

### 2022-01-12 writing/draft6/draft6.tex
draft

This paper presents a new linear dependent type system --- The Calculus of Linear Constructions (CLC). CLC extends the predicative CC$\omega$ with linear types. CC$\omega$ itself is an extension of CC with a cumulative hierarchy of type universes. We add an extra universe sort $L$ of linear types with cumulativity parallel to the sort $U$ of non-linear types. This ultimately cumulates in the \textit{linearity} theorem, stating that all resources are used exactly once.

@@

We further extend CLC with rules for defining inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). Standard Linear Logic connectives such a $\otimes$ and $\oplus$ are definable using this generalized mechanism in a straightforward way. Through examples in Section \ref{examples}, we show that CILC facilitates lightweight reasoning about stateful programs, protocol enforcement and in-place update of data structures.

@@

\noindent \textbf{\textit{Contributions}}:
Our contributions can be summarized as follows.
\begin{itemize}
  \item Fist, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types. The integration of linear types and dependent types allows CLC to directly and precisely reason about resource.
  \item Second, we study the meta-theory of CLC directly, showing that it satisfies the standard properties of confluence, validity and subject reduction.
  \item Next, we prove the linearity theorem, showing the tangible impact linear types have on the structure of terms. We also show that promotion and dereliction of linear logic are derivable as theorems for function types.
  \item Additionally, we observe that CLC is compatible with the predicative CC$\omega$, allowing us to construct a reduction preserving model of CLC in CC$\omega$, showing that CLC is consistent. We prove that all valid CC$\omega$ terms are valid CLC terms after adding minimal amounts of annotation.
  \item Furthermore, we extend CLC with rules to define inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). This is the first presentation of a generalized mechanism for defining inductive linear types that we are aware of.
  \item All major results have been formalized and proven correct in the Coq Proof Assistant with help from the Autosubst \cite{autosubst} library. To the best of our knowledge, our development is the first machine checked formalization of a linear dependent type theory.
  \item Finally, we give an implementation of CILC in OCaml. Algorithmic type checking employed by the typechecker streamlines the process of writing CILC programs.
\end{itemize}

\section{The Language of CLC}
\subsection{Syntax}
\begin{figure}[h]
  \caption{Syntax of CLC}
  \begin{grammar}
    <$i$> := 0 | 1 | 2 ... \phantom{* |} \hspace*{3em} universe levels

@@

A clear departure of our language from standard presentations of both linear type theory and dependent type theory is the presence of two function types: $(x :_s A) \rightarrow B$, $(x :_s A) \multimap B$. The reason for these function types is that we have built the ! exponential of linear logic implicitly into universe sorts. The sort annotation $s$ here records the universe sort of the function domain. The behavior of ! is difficult to account for even in simple linear type theories. Subtle issues arise if !! is not canonically isomorphic to !, which may invalidate the substitution lemma \cite{substitute}. By integrating the exponential directly into universe sorts, we have limited ! to only be canonically usable. This allows us to derive the substitution lemma, and construct a direct modeling of CLC in CC$\omega$ without requiring any additional machinery for manipulating exponential.

>>>>

This paper presents a new linear dependent type system --- the Calculus of Linear Constructions (CLC). CLC extends the predicative CC$\omega$ with linear types. CC$\omega$ itself is an extension of CC with a cumulative hierarchy of type universes. We add an extra universe sort $L$ of linear types with cumulativity parallel to the sort $U$ of non-linear types. This ultimately cumulates in the \textit{linearity} theorem, stating that all resources are used exactly once.

@@

We further extend CLC with rules for defining inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). Standard Linear Logic connectives such as $\otimes$ and $\oplus$ are definable using this generalized mechanism in a straightforward way. Through examples in Section \ref{examples}, we show that CILC facilitates lightweight reasoning about stateful programs, protocol enforcement and in-place update of data structures.

@@

\noindent \textbf{\textit{Contributions}}:
Our contributions can be summarized as follows.
\begin{itemize}
  \item First, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types. The integration of linear types and dependent types allows CLC to directly and precisely reason about resource.
  \item Second, we study the meta-theory of CLC directly, showing that it satisfies the standard properties of confluence, validity and subject reduction.
  \item Next, we prove the linearity theorem, showing the tangible impact linear types have on the structure of terms. We also show that promotion and dereliction of linear logic are derivable as theorems for function types.
  \item Additionally, we observe that CLC is compatible with the predicative CC$\omega$, allowing us to construct a reduction preserving model of CLC in CC$\omega$, showing that CLC is consistent. We prove that all valid CC$\omega$ terms are valid CLC terms after adding minimal amounts of annotation.
  \item Furthermore, we extend CLC with rules to define inductive linear types in the style of CIC, forming the Calculus of Inductive Linear Constructions (CILC). This is the first presentation of a generalized mechanism for defining inductive linear types that we are aware of.
  \item All major results have been formalized and proven correct in the Coq Proof Assistant with help from the Autosubst \cite{autosubst} library. To the best of our knowledge, our development is the first machine checked formalization of a linear dependent type theory.
  \item Finally, we give an implementation of CILC in OCaml. Algorithmic type checking employed by the typechecker streamlines the process of writing CILC programs.
\end{itemize}

\section{The Language of CLC}
\subsection{Syntax}
\begin{figure}[h]
  \caption{Syntax of CLC}
  \begin{grammar}
    <$i$> ::= 0 | 1 | 2 ... \phantom{* |} \hspace*{3em} universe levels

@@

A clear departure of our language from standard presentations of both linear type theory and dependent type theory is the presence of two function types: $(x :_s A) \rightarrow B$, $(x :_s A) \multimap B$. The reason for these function types is that we have built the ! exponential of linear logic implicitly into universe sorts. The sort annotation $s$ here records the universe sort of the function domain. The behavior of ! is difficult to account for even in simple linear type theories. Subtle issues arise if !! is not canonically isomorphic to !, which may invalidate the substitution lemma \cite{substitute}. By integrating the exponential directly into universe sorts, we have limited ! to only be canonically usable. This allows us to derive the substitution lemma and construct a direct modeling of CLC in CC$\omega$ without requiring any additional machinery for manipulating exponential.
